---
title: "Regression Models Course Project"
author: "Carlos Macasaet"
date: "24 August 2014"
output: pdf_document
---

## Executive Summary

Although manual transmission vehicles have a higher median fuel efficiency as
measured in miles-per-gallon (MPG), we cannot conclude that a manual
transmission is better for MPG. There are far better predictors of MPG. What we
can say is that in general, increasing the number of gears improves MPG, but for
manual transmission vehicles, that improvement is smaller.

## Is an automatic or manual transmission better for MPG?

Some exploratory data analysis shows that manual transmission vehicles have a
higher fuel efficiency in aggregate than automatic transmission vehicles
(figure 1). In addition, the initial analysis also suggests that fuel efficiency
has a linear relationship with the number of cylinders, displaplacement, weight,
and to a lesser extent, rear axle ratio (figures 1 and 2). However, a better
analysis is provided by a linear regression model.

```{r, echo=FALSE}
transmissionType <- as.factor( mtcars$am )
cylinders <- as.factor( mtcars$cyl )
engineConfiguration <- as.factor( mtcars$vs )
```

#### Analysis of Variance
I start by constructing thirteen increasingly sophisticated linear models
with the simplest one predicting MPG as a function of transmission type alone.

```{r, echo=FALSE}
anova( lm( mtcars$mpg ~ transmissionType ),
       lm( mtcars$mpg ~ transmissionType + cylinders ),
       lm( mtcars$mpg ~ transmissionType + cylinders + mtcars$disp ),
       lm( mtcars$mpg ~ transmissionType + cylinders + mtcars$disp +
             engineConfiguration ),
       lm( mtcars$mpg ~ transmissionType + cylinders + mtcars$disp +
             engineConfiguration + mtcars$carb ),
       lm( mtcars$mpg ~ transmissionType + cylinders + mtcars$disp +
             engineConfiguration + mtcars$carb + mtcars$gear ),
       lm( mtcars$mpg ~ transmissionType + cylinders + mtcars$disp +
             engineConfiguration + mtcars$carb + mtcars$gear + mtcars$wt ),
       lm( mtcars$mpg ~ transmissionType + cylinders + mtcars$disp +
             engineConfiguration + mtcars$carb + mtcars$gear + mtcars$wt +
             mtcars$drat ),
       lm( mtcars$mpg ~ transmissionType + cylinders * mtcars$disp +
             engineConfiguration + mtcars$carb + mtcars$gear + mtcars$wt +
             mtcars$drat ),
       lm( mtcars$mpg ~ transmissionType * mtcars$gear + cylinders +
             mtcars$disp + cylinders:mtcars$disp +
             cylinders:engineConfiguration + engineConfiguration + mtcars$carb +
             mtcars$wt * mtcars$drat ),
       lm( mtcars$mpg ~ transmissionType * mtcars$gear + cylinders +
             mtcars$disp + cylinders:mtcars$disp + engineConfiguration +
             cylinders:mtcars$wt + engineConfiguration + mtcars$carb +
             mtcars$wt * mtcars$drat ),
       lm( mtcars$mpg ~ transmissionType * mtcars$gear + cylinders +
             mtcars$disp + cylinders:mtcars$disp + engineConfiguration +
             cylinders:mtcars$wt + engineConfiguration + mtcars$carb +
             mtcars$wt * mtcars$drat + mtcars$hp ),
       lm( mtcars$mpg ~ transmissionType * mtcars$gear + cylinders +
             mtcars$disp + cylinders:mtcars$disp + engineConfiguration +
             cylinders:mtcars$wt + engineConfiguration + mtcars$carb +
             mtcars$wt * mtcars$drat +
             mtcars$hp + mtcars$qsec ) )
```

In the last two models (12 and 13) I introduce horsepower and quarter mile time
as variables. This gave me pause because like fuel efficiency they are outcomes
of vehicle design and therefore do not seem appropriate as predictors of MPG.
However, in case there were other aspects of vehicle design that significantly
contributed to these features but were not represented in the data, I decided to
test their effectiveness.

As it turned out adding horsepower and quarter mile time did not improve the
model significantly, so I chose to ignore them. In addition, adding an
interaction between number of cylinders and weight made the model worse. The
winner based on the analysis of variance is the tenth model which includes all
of the vehicle attributes plus interactions between transmission and number of
gears, number of cylinders and displacement, number of cylinders and engine
configuration, and weight and rear axle ratio.

### So is an automatic or manual transmission better for MPG?

```{r, echo=FALSE}
fit <-
  lm( mtcars$mpg ~ transmissionType * mtcars$gear + cylinders +
       mtcars$disp + cylinders:mtcars$disp +
       cylinders:engineConfiguration + engineConfiguration + mtcars$carb +
       mtcars$wt * mtcars$drat )
summary <- summary( fit )
summary$coefficients
```

According to the confidence levels of the various regressors, we cannot really
say. Transmission type has the 11th highest confidence level with less than 80%
confidence. This tells us that transmission type may have some impact on MPG,
but we do not have as high certainty about as we do others. The following table
lists the top ten predictors in their order of confidence level:

```{r, echo=FALSE}
coefficients <- summary$coefficients
summaryDf <-
  data.frame( coefficients[ 2:dim( coefficients )[ 1 ], ] )
summaryDf[ order( summaryDf[ , 4 ] ), 3:4 ][ 1:10, ]
```

Selecting only the features with 95% confidence or higher yields:

```{r, echo=FALSE}
fit <-
  lm( mtcars$mpg ~ mtcars$gear + cylinders + mtcars$disp +
        transmissionType:mtcars$gear + cylinders:mtcars$disp +
        cylinders:engineConfiguration )
summary( fit )$coefficients
```

### Diagnostics

Diagnostic analyses of the residuals show that this model is sound. There are no
outlying hat values or DF Betas. In addition, there is no pattern to the plot of
residuals to predictions (figure 3).

## Quantified MPG difference between automatic and manual transmissions

To answer this, we must interpret the coefficients of the linear regression. As
mentioned before, we do not have sufficiently high confidence of transmission as
a predictor of fuel efficiency on its own. However because of the strong
influence of its interaction with the number of gears, we can say that for
manual transmission vehicles, each additional gear reduces the fuel efficiency
by about two miles per gallon. But in general, each additional gear increases
the mileage by about five miles per gallon. So for automatic transmission
vehicles, each additional gear increases mileage by about five miles per gallon.
But for manual transmission vehicles, each gear only increases the mileage by
three miles per gallon.

## Appendix

```{r, fig.cap='Figure 1: impact of categorical features on fuel efficiency', echo=FALSE}
transmission <- as.factor( ifelse( mtcars$am == 0, 'automatic', 'manual' ) )
par( mfrow=c( 2, 3 ), oma=c( 1, 1, 1, 1 ) )
boxplot( mtcars$mpg ~ transmission, xlab='Transmission Type', ylab='MPG' )
boxplot( mtcars$mpg ~ as.factor( mtcars$cyl ), xlab='Cylinders', ylab='MPG' )
boxplot( mtcars$mpg ~ as.factor( ifelse( mtcars$vs == 0, 'straight', 'v' ) ),
         xlab='Engine Configuration', ylab='MPG' )
boxplot( mtcars$mpg ~ as.factor( mtcars$gear ), xlab='Number of Gears',
         ylab='MPG' )
boxplot( mtcars$mpg ~ as.factor( mtcars$carb ), xlab='Number of Carburetors',
         ylab='MPG' )
title( 'Figure 1: impact of categorical features on fuel efficiency',
       outer=TRUE)
```

```{r, fig.cap='Figure 2: impact of continuous features on fuel efficiency', echo=FALSE}
par( mfrow=c( 2, 2 ), oma=c( 1, 1, 1, 1 ) )
plot( mtcars$mpg ~ mtcars$disp, xlab='Displacement', ylab='MPG' )
plot( mtcars$mpg ~ mtcars$wt, xlab='Weight', ylab='MPG' )
plot( mtcars$mpg ~ mtcars$drat, xlab='Rear Axle Ratio', ylab='MPG' )
title( 'Figure 2: impact of continuous features on fuel efficiency',
       outer=TRUE )
```

```{r, fig.cap='Figure 3: Residual Plots', echo=FALSE}
par( mfrow=c( 2, 2 ), oma=c( 1, 1, 1, 1 ) )
plot( hatvalues( fit ), xlab='Observation', ylab='Hat Value' )
plot( dfbetas( fit )[ , 2 ], xlab='Observation', ylab='DF Beta' )
plot( predict( fit ), resid( fit ), xlab='Predicted MPG', ylab='Residual' )
plot( resid( fit ) / ( 1 - hatvalues( fit ) ), xlab='Observation',
      ylab='PRESS Residual' )
title( 'Figure 3: Residual Plots', outer=TRUE )
```